---
title: "Exploratory Data Analysis of Incomes and Expenses for United States Colleges"
format: html
editor: visual
---

### Set Up

Import data and necessary libraries.

```{r}
library(tidyverse)
library(vtable)
library(ggplot2)
library(GGally)
library(gridExtra)
library(ggpubr)

df <- readRDS('./data/college_expenses_and_enrollment.Rdata')
```

### Variable Review

```{r}
vtable::vtable(df)
```

Non-income or expense variables include unique college ID, state abbreviation, college type, institution name, total enrollment, full time enrollment, and undergraduate enrollment. The state and college type factors might be helpful for grouping. The institution name for labeling particular values. Total enrollment is interesting because the minimum value is 9. There was a college with 9 students? Could be an error. A 0 value minimum for full time and undergraduate enrollment makes a little more sense to me, perhaps there are some community colleges, etc. with only part-time students or perhaps college only accepts graduate students, I'm not sure, but also something to investigate. Per the assignment note, most values will indicate the size and the enrollment values can be used to calculate per capita numbers for comparison. It may make the most sense to look at four year schools with full-time undergraduate students for per capita comparisons. I would like to know what year this data set is from. Of note, both college_type and state data type is factor. I think it makes sense to probably change to character.

Income variables include total, tuition, federal support, state support, local support, private support, and sales. I'm assuming income is gross and not net. Some colleges have really low incomes (\$6,067?) but I don't have enough domain knowledge to know if that is reasonable yet. Also what school has a \$0 income from tuition? Do they have no students? I'm also not sure why there would be a negative value for private support or sales income. Income might actually be net (revenues - expenses). I attempted to find more information about the variable descriptions via the \[IPEDS Survey Components\](https://nces.ed.gov/ipeds/use-the-data/survey-components/2/finance) and \[IPEDS Compare Institutions Variable Tree\](https://nces.ed.gov/ipeds/datacenter/SelectVariables.aspx?stepId=1). To be honest, I'm not quite sure about the data that I'm working with as it was difficult to match variables in the file provided to the site - good real world experience - and I'm going to have to make some assumptions. Biggest call outs so far are (1) finance variables are nested under the type of institution and I'm not quite sure how they are derived in this data set so this may lead to some unexpected variation, (2) there is a difference between Fall Enrollment and 12-Month Enrollment, I'm not quite sure what values are provided in the data set.

Expense variables include total, research, public service, student services, instruction, academic support, and other. There are negative values in public service and other, not sure why yet. I wasn't sure what all of these categories mean, so using prior links and this \[Postsecondary Institution Expenses article\](https://nces.ed.gov/programs/coe/indicator/cue/postsecondary-institution-expense#fn6) as best reference. Research is any activity that produces research. Public service is any activity that provides non-instructional services to benefit people outside of the institution (community service, cooperative extension services, public broadcasting services). Student services is any activity that supports student well-being and development outside of formal instruction (admissions, registrar activities, dependent intercollegiate athletics and health services). Instruction is any activity that provides instruction (credit and non-credit), may also include research and public service activities that are not separately budgeted. Academic support is any activity that supports the institution's mission of instruction, research, and public service. Other expenses is a catch-all for items such as bad debts, income taxes, liability obligations, or other operational expenses not associated with the primary mission of the institution, such as a major federally funded research and development center. The IPEDS Expenses article groups expenses as follows:\
1. Independent operations and other\
2. Hospital services\
3. Net grant aid to students\
4. Auxiliary enterprises\
5. Academic support, student services, and institutional support\
6. Research and public service\
7. Instruction

Assuming income means revenue, there is also an IPEDS \[Postsecondary Institution Revenues article\](https://nces.ed.gov/programs/coe/indicator/cud) to help inform what these columns mean. I get the gist of tuition, federal, state, and local incomes. Private income includes any gifts, grants, and contracts from non-governmental entities. Sales income includes revenues from sales of goods and services such as publications, testing, cosmetology, and hospital. I'm not sure about the data provided, but in the article, private grants and contracts are included in the local income instead of private income. In addition, income is not comparable across types of institutions because they may treat income differently. For example, Pell Grants are usually part of Federal funding for public institutions but usually part of tuition funding for private institutions. The article also groups revenue as follows:\
1. Tuition and fees\
2. Investments\
3. Government\
3. Auxiliary enterprises\
4. Other

Whew! Lots of questions remaining, but I think I have a better understanding of what some of these variables mean and areas to proceed with caution. And it may make sense to group some of the columns similar to the articles referenced. Also just re-read assignment. The important note uses income interchangeably with revenue, so that question is answered.

### Data Preparation

To prepare the data, I renamed the columns for clarification and consistency, then completed the following checks to handle missing values and ensure data seems reasonable. I want to do some initial exploratory data analysis before deciding how to derive per capita values.

```{r}
# Rename columns and convert state and college type to character.
df <- df %>%
  rename(college_id = UNITID,
         state = STABBR,
         institution = Institution.Name,
         total_enrollment = Total.Enrollment,
         full_time_enrollment = Full.Time.Enrollment,
         undergraduate_enrollment = Undergraduate.Enrollment,
         college_type = type,
         total_income = Total.Income,
         tuition_income = Tuition,
         federal_income = Federal,
         state_income = State,
         local_income = Local,
         private_income = Private,
         sales_income = Sales,
         total_expense = Total.Expenses,
         research_expense = Research,
         public_service_expense = Public.Service,
         student_service_expense = Student.Services,
         instruction_expense = Instruction,
         academic_support_expense = Academic.Support,
         other_expense = Other) %>%
  mutate(across(where(is.factor), as.character))

# Update vtable output to match.
vtable::vtable(df)
```

1.  Check for missing values and resolve.

```{r}
# Summarize number of missing values per column where they exist.
df %>%
  summarize_if(anyNA, funs(sum(is.na(.))))

# There are 426 missing institution names and 516 missing values for each enrollment columm. Although it decreases the data set, I can't think of a reasonable reason why an institution would not have a name or any total enrollment so I think it makes sense to remove these rows from the main data set. Since the number of missing values for each enrollment column is the same, I imagine it is the same rows for each column. If not, it is reasonable that a college may not have any full-time students or undergraduate students, or the total enrollment value could be missing due to an error. Check that all missing enrollment values belong to the same rows.

df %>%
  filter(is.na(total_enrollment)) %>%
  select(total_enrollment,
         full_time_enrollment,
         undergraduate_enrollment) %>%
  summarize_all(funs(sum(is.na(.))))

# Confirmed. Separate out missing values from main data set. 

main_df <- df %>%
  drop_na()

na_df <- df %>%
  filter(if_any(everything(), ~ is.na(.x)))

# Reviewing NAs, based on the institution names, many of these are system or district administration offices, so it makes sense to have no enrollment. I still think it makes sense to exclude these institutions because I would like to focus on institutions with students and be able to compare per capita values.
```

2.  Check that unique values make sense. All values need to be unique for college_id, approximately 50 unique values for state (unless territories are included, if yes, rename column to clarify), and college types need to make sense.

```{r}
# Select columns and count unique values.
main_df %>%
  select(college_id, state, college_type) %>%
  summarize_all(funs(n_distinct(.)))

# The number of unique college_id values is equal to the number of observations. There is more than 50 states so reviewing values.

states_df <- main_df %>%
  distinct(state) %>%
  arrange(state)

# Reviewing against the IPEDS variable tree, the additional nine values are: District of Columbia (DC), American Samoa (AS), Federated States of Micronesia (FM), Guam (GU), Marshall Islands (MH), Northern Marianas (MP), Palau (PW), Puerto Rico (PR), and the Virgin Islands (VI). Renaming column to clarify

main_df <- main_df %>%
  rename(state_or_territory = state)

# Review unique college types.
main_df %>%
  distinct(college_type)

# All college types makes sense (NP = non-profit). Types vary by 2/4 year, public/private, and for/non profit.
```

3.  For income and expense values: no negative values, a zero value may also be suspect (a school does not spend any funding on instruction?) but will keep in for now and evaluate during exploratory data analysis. In addition, the sum of income or expense subcategories should not be more than the total.

```{r}
# Review income and expense columns with negative observations. I could be missing something, but I don't think it make sense to have negative values in any of these columns.
vtable::vtable(main_df)

# There are negative values for private_income, sales_income, and public service_expense. Investigate further to better understand and inform next steps.
main_df %>%
  filter(if_any(everything(), ~ .x < 0))
  
# There are only 8 rows with a negative value in any column. It seems like these might be errors. Since it is very few rows, I think it makes more sense to remove rows rather than set to zero or remove the negative sign. 

neg_df <- main_df %>%
  filter(if_any(everything(), ~ .x < 0))

main_df <- main_df %>%
  filter(if_all(everything(), ~ .x >= 0))

# Check for observations where subcategory incomes are greater than the total income. There may be some subcategory incomes not listed (investments?)
main_df %>%
  select(contains("income")) %>%
  mutate(calc_income = tuition_income +
           federal_income +
           state_income +
           local_income +
           private_income +
           sales_income,
         income_error = if_else(total_income == calc_income, 0, 1)) %>%
  filter(income_error == 1)

# No observations where the calculated income is greater than the total income. Check the same for expenses.
main_df %>%
  select(contains("expense")) %>%
  mutate(calc_expense = research_expense +
           public_service_expense +
           student_service_expense +
           instruction_expense +
           academic_support_expense +
           other_expense,
         expense_error = if_else(total_expense == calc_expense, 0, 1)) %>%
  filter(expense_error == 1)

# No expense errors where calculated expense is greater than total expense.
```

4.  For enrollment values: apparently some colleges can be really small (see \[Deep Springs College, CO\](https://en.wikipedia.org/wiki/Deep_Springs_College), so a minimum of 9 might be reasonable, but unlikely. Check that full-time or undergraduate enrollment does not exceed total enrollment. Evaluate outliers during exploratory data analysis.

```{r}
# Check that enrollment values make sense.
main_df %>%
  select(contains("enrollment")) %>%
  mutate(full_time_error = if_else(total_enrollment >= full_time_enrollment, 0, 1),
         undergrad_error = if_else(total_enrollment >= undergraduate_enrollment, 0, 1)) %>%
  filter(full_time_error == 1 | undergrad_error == 1)

# No errors where the number of full-time or undegraduate students is greater than the total enrollment.
```

### Exploratory Data Analysis - Univariate Non-Graphical

```{r}
# Summarize main data.
vtable(main_df, lush = TRUE)
sumtable(main_df)
```

No missing values, as expected. There is one state or territory that is no longer represented. Will look at graphs for clearer picture of income and expense values. Percentage breakdown of college types is interesting. The majority of colleges are private non-profit 4-year schools (36.5%), followed by public 2-year (22.7%), public 4-year (17%), for-profit 2-year (13.4%), for-profit 4-year (7.2%), and lastly private non-profit 2-year (3.2%). Missed earlier that not all the institution names are unique. I'm guessing there might be multiple institutions with the same name, sub-entities, or duplicates. Need to further investigate and potentially remove.

```{r}
# Review duplicate institutions
duplicates_df <- main_df %>%
  group_by(institution) %>%
  filter(n() > 1) %>%
  arrange(institution)

# Check for any exact matches.
duplicates_df %>%
  group_by_all() %>%
  filter(n() > 1) %>%
  ungroup()

# There are no exact row matches.
```

Duplicate institution names are due to several reasons. I also reviewed the definition of UNITID (college_id). It is "Unique identification number assigned to postsecondary institutions surveyed through the Integrated Postsecondary Education Data System (IPEDS). Also referred to as UNITID or IPEDS ID." The main ones seem to be as follows:

1.  Each campus belonging to the institution has a unique college ID. For example Academy of Interactive Entertainment has a unique ID for their Washington state campus and Louisiana campus. It may be in the same state as well, such as American College of Healthcare and Technology with two California campuses.
2.  There are separate institutions with the same name, such as the two Anderson colleges.

I decided to leave in duplicate names and continue using the college_id as the identifier. I think it is likely some separate college campuses are essentially separate, if affiliate institutions.

### Exploratory Data Analysis - Univariate Graphical

First, look at the distributions for continuous variables.

```{r}
# Create list of continuous target variables.
target_variables <- main_df %>%
  select(matches("income|expense|enrollment")) %>%
  colnames()

# Visualize historgrams of target variables.
for (variable in target_variables) {
  print(ggplot(main_df, aes_string(variable)) + 
          geom_histogram())
}
```

All continuous variables are heavily skewed right. To me, this mostly makes sense. It's less likely to have colleges with high sources of income, expenses, or enrollments.

Second, look at the distribution of categorical variables.

```{r}
# college type
ggplot(main_df) +
  geom_bar(aes(x = college_type, y = after_stat(count/sum(count)))) +
  coord_flip() + 
  theme_classic() +
  scale_y_continuous(labels = scales::percent) +
  ylab("percentage")

# state or territory
ggplot(main_df) +
  geom_bar(aes(x = state_or_territory, y = after_stat(count/sum(count)))) +
  coord_flip() + 
  theme_classic() +
  scale_y_continuous(labels = scales::percent) +
  ylab("percentage")
```

I wonder why the private non-profit 4-year colleges have the highest number of schools and the private non-profit 2-year colleges have the lowest number of schools. This is also a flipped relationship compared to for-profit and public schools where there are more 2-year colleges than 4-year colleges. Per the articles mentioned prior, may make sense to group by public, private for-profit, and private non-profit. In hindsight, that's probably why the variable was original a 2x3 factor. The state_or_territory distribution is squished, but California clearly has the highest percentage of institutions. It may make sense to summarize by \[region\](https://nces.ed.gov/ipeds/Search/ViewTable?tableId=10788#:\~:text=Utah%2C%20and%20Wyoming.-,The%20Far%20West%20region%20includes%20Alaska%2C%20California%2C%20Hawaii%2C,Nevada%2C%20Oregon%2C%20and%20Washington.).

### Exploratory Data Analysis - Multivariate Non-Graphical

```{r}
# Group by college_type
main_df %>%
  sumtable(group = 'college_type')
```

Enrollment highly varies by college_type. For example, the for-profit 2-year program has fairly close average total, undergraduate, and full-time enrollments. On average, there is 360 students per program, all of which are undergraduate and most are full time. Compare these data to the public 4-year program where the average total enrollment is 12,140 students per program, however, the average number full-time students is 8,635 per program, and 10,127 students on average are undergraduates. Standard deviations are very high for all categories, so I think this is where college size comes into play and calculating per capita values are important.

### Exploratory Data Analysis - Multivariate Graphical

There's a lot of directions to go with exploratory data analysis. I found the IPEDS visuals from the articles linked earlier inspiring, so I would like to explore something similar. Reviewing, it sounds like "per capita" is calculated as per full-time student plus part-time equivalent. I think it makes the most sense to follow a similar calculation - I don't know if graduate versus undergraduate tuition costs are that different, or at least I think full-time versus part-time is more important. In addition this "per capita" calculation, I decided to group the college types and financial categories. Due to time constraints, I focused on income.

```{r}
# Calculate financial values per full time equivalent (FTE) student. Assuming that a part-time student exactly half of a full-time student, and rounding up to provide a more conservative estimate.
per_cap_df <- main_df %>%
  mutate(
    government_income = federal_income + state_income + local_income,
    other_income = private_income + sales_income,
    re_pub_serv_expense = research_expense + public_service_expense,
    stu_ser_aca_sup_expense = student_service_expense + academic_support_expense,
    fte = round(
    full_time_enrollment + ((total_enrollment - full_time_enrollment) / 2),
    0),
    across((contains("income") | contains("expense")),  ~ .x / fte),
    program_type = case_when(
      grepl("For-Profit", college_type) ~ "Private for-profit",
      grepl("Private NP", college_type) ~ "Private nonprofit",
      grepl("Public", college_type) ~ "Public",
      TRUE ~ "Error"
      ),
    program_year = case_when(grepl("2-Year", college_type) ~ 2,
                             grepl("4-Year", college_type) ~ 4,
                             TRUE ~ 0)
    )

# Check if situations where income does not add up as test. This was happening due to slight differences with how floating point numbers are calculated, numbers are the same (see work below).

per_cap_df %>%
  select(contains("income")) %>%
  mutate(calc_income_1 = tuition_income +
           federal_income +
           state_income +
           local_income +
           private_income +
           sales_income,
         calc_income_2 = (tuition_income + 
                          government_income +
                          other_income),
         income_error_1 = if_else(
           round(total_income, 0) == round(calc_income_1, 0), 0, 1),
         income_error_2 = if_else(
           round(total_income, 0) == round(calc_income_2, 0), 0, 1)) %>%
  select(total_income, calc_income_1, calc_income_2, income_error_1, income_error_2) %>%
    filter((income_error_1 == 1) | (income_error_2 == 1))

# Review high-level financial graphs by program type. Note: Tried additional variables but too crowded.

fin_df <- per_cap_df %>%
  select(program_type,
         total_income,
         total_expense)

GGally::ggpairs(fin_df)

# Prepare income graph. Pivot to long form to get funding source category.
income_df <- per_cap_df %>%
  group_by(program_type) %>%
  select(program_type, tuition_income, government_income, other_income) %>%
  pivot_longer(cols = ends_with("income"), names_to = "income_source", values_to = "amount")

income_df <- income_df %>%
  group_by(program_type, income_source) %>%
  summarize(total_amount = sum(amount)) %>%
  group_by(program_type) %>%
  mutate(percent = total_amount / sum(total_amount))
  
```

Graph percent distribution of total income by control of institution and funding source.

```{r}
# Prepare data.
income_df <- main_df %>%
  mutate(government_income = federal_income + state_income + local_income,
         other_income = private_income + sales_income,
         program_type = case_when(
           grepl("For-Profit", college_type) ~ "Private for-profit",
           grepl("Private NP", college_type) ~ "Private nonprofit",
           grepl("Public", college_type) ~ "Public",
           TRUE ~ "Error")) %>%
  select(program_type, tuition_income, government_income, other_income) %>%
  pivot_longer(cols = ends_with("income"), names_to = "source", values_to = "amount") %>%
  group_by(program_type, source) %>%
  summarize(total_amount = sum(amount)) %>%
  group_by(program_type) %>%
  mutate(percent = total_amount / sum(total_amount))
  
# Graph.
ggplot(income_df, aes(x = program_type, y = total_amount, fill = source)) + 
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")),
                     position=position_fill(vjust=0.5), color="black") +
  labs(title = "Percent Distribution of Total Income \nby Control of Institution and Funding Source") +
  xlab(label = "Control of Institution") +
  ylab(label = "Percent of Income") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(name = "Funding Source",
                      labels=c("Government", "Private and Sales", "Tuition and Fees")) +
  theme_minimal()
```

Keeping in mind that funding for different institutional controls cannot be compared because they label funding sources differently...

Over three-fourths of private for-profit funding is from tuition and fees with almost a fifth from the government and only a little from private and sales. Two-thirds of private nonprofit funding is from private and sales sources, less than one-third is from tuition and fees, and only a little from the government. Funding for public institutions is approximately two-fifths from government sources, one-third from private and sales sources, and less than one-fourth from tuition and fees. I think this is helpful to see where funding comes from for each type of institutional control.

Graph tuition and fee income per full-time equivalent student by control of institution.

```{r}
# Prepare data. 
tuition_df <- main_df %>%
  select(tuition_income, total_enrollment, college_type, full_time_enrollment) %>%
  mutate(fte = round(
    full_time_enrollment + ((total_enrollment - full_time_enrollment) / 2),
    0),
    tuition_per_fte = (tuition_income / fte),
    program_type = case_when(
      grepl("For-Profit", college_type) ~ "Private for-profit",
      grepl("Private NP", college_type) ~ "Private nonprofit",
      grepl("Public", college_type) ~ "Public",
      TRUE ~ "Error")) %>%
  select(tuition_per_fte, program_type)

# Figure out color.
library("RColorBrewer")
display.brewer.pal(9, "Blues")
brewer.pal(9, "Blues")

# Check shape of tuition per full-time equivalent student to see if it makes sense to take mean or median.
ggplot(tuition_df, aes(tuition_per_fte)) +
  geom_histogram()

# Data is heavily skewed right, so using the median to determine average tuition per full time equivalent student by control of institution.
tuition_df %>%
  group_by(program_type) %>%
  summarize(median_per_fte = median(tuition_per_fte)) %>%
ggplot() +
  geom_col(aes(x = program_type, y = median_per_fte), fill = "#4292C6") +
  labs(title = "Tuition and Fees Income per Full-Time Equivalent (FTE) Student \n by Control of Institution",
       subtitle = "(Averaged by Median)") +
 xlab(label = "Control of Institution") +
 ylab(label = "Tuition and Fees Income") +
  geom_text(aes(x = program_type, y = median_per_fte, label = scales::dollar(round(median_per_fte, 0))),
            color = "white",
            vjust = 1.5) +
  scale_y_continuous(labels=scales::dollar_format()) +
  scale_color_brewer() +
  theme_minimal()
```

The average tuition and fees per full-time equivalent student is highest for private nonprofits, closely followed by private for-profits, and much lower for public institutions. This may be because private institutions treat some government funding as tuition income.

Graph government income per full-time equivalent student by funding source and control of the institution.

```{r}
# Prepare data. 
gov_df <- main_df %>%
  select(federal_income, state_income, local_income, total_enrollment, college_type, full_time_enrollment) %>%
  mutate(fte = round(
    full_time_enrollment + ((total_enrollment - full_time_enrollment) / 2),
    0),
    across(contains("income"),  ~ .x / fte),
    program_type = case_when(
      grepl("For-Profit", college_type) ~ "Private for-profit",
      grepl("Private NP", college_type) ~ "Private nonprofit",
      grepl("Public", college_type) ~ "Public",
      TRUE ~ "Error")) %>%
  select(federal_income, state_income, local_income, program_type) %>%
  mutate(state_and_local_income = state_income + local_income)

# Check shape of government incomes per full-time equivalent student to see if it makes sense to take mean or median.
gov_df %>%
  filter(program_type == "Private nonprofit") %>%
ggplot(aes(x = federal_income)) +
  geom_histogram()

# Create list of government income variables.
gov_variables <- gov_df %>%
  select(matches("income")) %>%
  colnames()

# Visualize historgrams of target variables.
for (variable in gov_variables) {
  print(ggplot(gov_df, aes_string(variable)) + 
          geom_histogram())
}

GGally::ggpairs(gov_df)

# Played around with different program_types and I think it still makes sense to take the median.Most data is heavily skewed right.
fed_plot <- gov_df %>%
  group_by(program_type) %>%
  summarize(median_per_fte = median(federal_income)) %>%
ggplot() +
  geom_col(aes(x = program_type, y = median_per_fte), fill = "#4292C6") +
  labs(title = "Federal Income per Full-Time Equivalent (FTE) Student \n by Control of Institution",
       subtitle = "(Averaged by Median)") +
 xlab(label = "Control of Institution") +
 ylab(label = "Federal Income") +
  geom_text(aes(x = program_type, y = median_per_fte, label = scales::dollar(round(median_per_fte, 0))),
            color = "white",
            vjust = 1.5) +
  scale_y_continuous(labels=scales::dollar_format()) +
  theme_minimal()

state_and_local_plot <- gov_df %>%
  group_by(program_type) %>%
  summarize(median_per_fte = median(state_and_local_income)) %>%
ggplot() +
  geom_col(aes(x = program_type, y = median_per_fte), fill = "#4292C6") +
  labs(title = "State and Local Income per Full-Time Equivalent (FTE) Student \n by Control of Institution",
       subtitle = "(Averaged by Median)") +
 xlab(label = "Control of Institution") +
 ylab(label = "State and Local Income") +
  geom_text(aes(x = program_type, y = median_per_fte, label = scales::dollar(round(median_per_fte, 0))),
            color = "black",
            vjust = -0.2) +
  scale_y_continuous(labels=scales::dollar_format()) +
  theme_minimal()

ggarrange(fed_plot, state_and_local_plot)

# After playing around with different libraries, I think it may work best to redo graphs with a facet grid.

gov_df %>%
  group_by(program_type) %>%
  summarize(avg_fed = median(federal_income), 
            avg_state_local = median(state_and_local_income)) %>%
  pivot_longer(contains("avg"), names_to = "source", values_to = "avg_amount") %>%
  mutate(source = case_when(source == "avg_fed" ~ "Federal",
                            source == "avg_state_local" ~ "State and Local",
                            TRUE ~ source)) %>%
  ggplot(aes(program_type, avg_amount)) +
  geom_col(fill = "#4292C6") +
  labs(title = "Government Income per Full-Time Equivalent (FTE) Student \nby Control of Institution",
       subtitle = "(Averaged by Median)") +
 xlab(label = "Control of Institution") +
 ylab(label = "Income") +
  geom_text(aes(x = program_type, 
                y = avg_amount, 
                label = scales::dollar(round(avg_amount, 0))),
            color = "#333333",
            vjust = -0.2,
            size = 3) +
  scale_y_continuous(labels=scales::dollar_format()) +
  facet_wrap(vars(source)) +
  theme_minimal()

```

Public institutions get the most government funding per student with significantly more state and local than federal funding. All institutions receive some type of federal funding, but private for-profit institutions do not appear to receive any state and local funding, and private nonprofit institutions receive very little state and local funding.

### Interesting Relationship Visual

It was a helpful learning exercise to try and create similar income graphs like on the IPEDS site, and I felt like all were somewhat confusing or misleading because each institutional control type really can't be compared to each other because they label funding sources differently. For an interesting relationship visual, I decided to create a funding source by control of institution that tried to separate out these categories better.

```{r}
income_graph <- main_df %>%
  select(federal_income, state_income, local_income, tuition_income, private_income, sales_income, total_enrollment, college_type, full_time_enrollment) %>%
  mutate(fte = round(
    full_time_enrollment + ((total_enrollment - full_time_enrollment) / 2),
    0),
    across(contains("income"),  ~ .x / fte),
    program_type = case_when(
      grepl("For-Profit", college_type) ~ "Private for-profit",
      grepl("Private NP", college_type) ~ "Private nonprofit",
      grepl("Public", college_type) ~ "Public",
      TRUE ~ "Error")) %>%
  select(federal_income, state_income, local_income, tuition_income, private_income, sales_income, program_type) %>%
  mutate(state_and_local_income = state_income + local_income,
         private_and_sales_income = private_income + sales_income) %>%
  group_by(program_type) %>%
  summarize("Federal" = median(federal_income), 
            "State and Local" = median(state_and_local_income),
            "Tuition and Fees" = median(tuition_income),
            "Private and Sales" = median(private_and_sales_income)) %>%
  pivot_longer(c("Federal", "State and Local", "Tuition and Fees",
                 "Private and Sales"), names_to = "source", values_to = "avg_amount") %>%
  group_by(program_type, source) %>%
  summarize(total = sum(avg_amount)) %>%
  group_by(program_type) %>%
  mutate(percent = total / sum(total)) %>%
  ggplot(aes(source, total, fill = factor(program_type))) +
  geom_col() +
  labs(title = "Median Income per Full-Time Equivalent (FTE) Student by Institution Type") +
  xlab(label = "Funding Source") +
  ylab(label = "Income") +
  geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")),
            vjust = -0.2, color="black", size = 3) +
  scale_y_continuous(labels=scales::dollar_format()) +
  facet_wrap(vars(program_type)) +
  scale_x_discrete(guide = guide_axis(angle = 50)) +
  scale_fill_brewer() +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        legend.position="none")

income_graph

ggsave("income_graph.png")
```

I think this visualization does a better job of separating out income for each institution, however, there is still room for improvement. I'm not sure yet what is the best way to handle the duplicative x labels (tried a few methods and underneath slanted seems to be best), and I'd like to add clearer breaks between the graphs.
